Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-05 exp1/iswc2013.tex
Introduction
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@504 43ef1190-90fd-4dd7-a769-d0121f02b80e

\title{Introducing Statistical Design of Experiments to SPARQL
  Endpoint Evaluation}
\author{Kjetil Kjernsmo\inst{1}}
%\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
%\institute{Department of Informatics,
%Postboks 1080 Blindern,
%0316 Oslo, Norway
%\email{kjekje@ifi.uio.no}}

%\maketitle

@@

@@

>>>>

\title{Introducing Statistical Design of Experiments to SPARQL
  Endpoint Evaluation}
%\author{Kjetil Kjernsmo\inst{1}}
%\titlerunning{Introducing Design of Experiments to SPARQL Evaluation}
%\institute{Department of Informatics,
%Postboks 1080 Blindern,
%0316 Oslo, Norway
%\email{kjekje@ifi.uio.no}}

@@

\maketitle

@@

The practice of benchmarking is widespread, both in industry to select
the most appropriate implementation amongst several choices, and in
academia as means to verify or refute an assertion on the performance
of a certain practice or system. While we could contribute to the
industrial cases, our focus is on the latter. A benchmark typically
consists of several microbenchmarks, which are procedures for placing
loads on the system to measure its performance as a function of the
load. For SPARQL benchmarks, usually the performance is measured as a
the number of queries the SPARQL implementation is able to answer in a
certain time (the throughput) or the time required for it to respond
to a query.

However, there are many of problems with this approach when used to
evaluate assertions about the overall performance of a system or
several systems: 
\begin{itemize}
\item As the complexity increases, the number of parameters that
one may wish to test increases dramatically.
\item Nothing compels anyone to find microbenchmarks that
  can refute an assertion, and there is no structured approach to
  investigate whether a benchmark is flawed.
\item There is no meaningful summary of the performance as a whole, so
  that an assertion about performance cannot be refuted as an
  optimization may have dentrimental side effects that cannot be
  identified, or cannot be compared to possibly positive effects in a
  different microbenchmark.
\end{itemize}

To address some of these problems, one may try to standardize
benchmarks, to eliminate some parameters and identify others as
believed to be important. Also, social pressure should then prompt
developers to use a standard benchmark so that assertions may be
refuted.

While this is an improvement which is possibly sufficient for
industrial uses, is it still flawed as empirical
research, as it cannot capture the full diversity of the problem: For
example, if a vendor claims that their implementation is better if
used on a highly capable hardware platform, then the standardization
of a hardware platform cannot test this assertion. Moreover, to test
this assertion, one would need a summary technique that can identify
the interaction between implementation and hardware platform and
possibly other parameters.

One important aspect of science is the structured approach to falsify
hypotheses, and thus is important that assertions about performance is
formulated as hypotheses that can be falsified.

Thus, attacking the complexity problem by eliminating parameters we
think influence the outcome for simplicity is indefensible as
scientific practice. We must instead tackle this problem by enhancing
our capability of testing complex systems and to identify situations
where our assumptions are flawed. Such constraints can only be made
when we have very good reason to suspect that a certain parameter does
not have influence.

In this paper, we employ techniques from the well-established field in
statistics of Design of Experiments to point out a direction that
promises to address these problems. First, we design a simple
8-parameter experiment, and then we demonstrate how a full
factorial experiment can meaningfully provide a summary of the
statistics.

Then, we demonstrate how a fractional factorial experiment can be used
to maintain a structured approach to the problem while reducing the
number of runs the experiment needs, and we discuss the tradeoffs
involved. We then continue to discuss techniques that shows our
simplistic experiment to be flawed, yet scientifically better than the
practice of benchmarking. Finally, we outline a roadmap to establish
the use of these techniques in SPARQL endpoint evaluations.
